---
title: "redeemR Filter2 QC Report"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
params:
  input_rds: null
  sample_name: null
  thr: null
  raw_name: "RawGenotypes.Sensitive.StrandBalance"
  top_n_variants: 20
  consensus_n: 25
  output_metrics_tsv: null
  output_report_rds: null
---

```{r}
stopifnot(!is.null(params$input_rds), file.exists(params$input_rds))

suppressWarnings(suppressPackageStartupMessages({
  library(redeemR)
  library(dplyr)
  library(ggplot2)
  library(knitr)
  library(stringr)
  library(Matrix.utils)
}))

hotcall_variants <- c("310_T_C", "9979_G_A", "3109_T_C")

sample_name <- params$sample_name
if (is.null(sample_name) || !nzchar(sample_name)) {
  sample_name <- sub("\\..*$", "", basename(params$input_rds))
}

redeem_obj <- readRDS(params$input_rds)

thr <- params$thr
if (is.null(thr) || !nzchar(thr)) {
  thr <- unname(redeem_obj@para["Threhold"])
}

final_dir <- redeem_obj@attr$path
input_rds_abs <- normalizePath(params$input_rds, mustWork = TRUE)
path_parts <- strsplit(input_rds_abs, "/")[[1]]
filter2_idx <- which(path_parts == "filter2")
if (length(filter2_idx) == 1L && filter2_idx > 2L && path_parts[filter2_idx - 1L] == "preprocessing") {
  project_root <- paste(path_parts[seq_len(filter2_idx - 2L)], collapse = "/")
  if (!nzchar(project_root)) {
    project_root <- "/"
  }
  project_root <- normalizePath(project_root, mustWork = TRUE)
} else {
  project_root <- normalizePath(dirname(dirname(dirname(dirname(input_rds_abs)))), mustWork = TRUE)
}
if (grepl("^/", final_dir)) {
  final_dir_abs <- final_dir
} else {
  final_dir_abs <- file.path(project_root, final_dir)
  if (!file.exists(final_dir_abs)) {
    final_dir_alt <- sub("^Analysis_redeemV/", "preprocessing/redeemV/", final_dir)
    final_dir_abs <- file.path(project_root, final_dir_alt)
  }
}
redeem_obj@attr$path <- final_dir_abs
sample_dir <- dirname(dirname(final_dir_abs))
redeemv_qc_png <- file.path(sample_dir, paste0(sample_name, ".QCplot.png"))
redeemv_qc_png_local <- file.path(getwd(), paste0(sample_name, ".redeemV_qc.png"))
if (file.exists(redeemv_qc_png)) {
  tmp_copy_ok <- file.copy(redeemv_qc_png, redeemv_qc_png_local, overwrite = TRUE)
  invisible(tmp_copy_ok)
}

capture_show <- function(x) {
  paste(capture.output(show(x)), collapse = "\n")
}

safe_head <- function(df, n = 10) {
  if (is.null(df) || !NROW(df)) {
    return(data.frame())
  }
  utils::head(as.data.frame(df), n)
}

safe_plot <- function(expr) {
  withCallingHandlers(
    tryCatch(expr, error = function(e) e),
    message = function(m) {
      invokeRestart("muffleMessage")
    },
    warning = function(w) {
      invokeRestart("muffleWarning")
    }
  )
}

quiet_eval <- function(expr) {
  withCallingHandlers(
    expr,
    message = function(m) {
      invokeRestart("muffleMessage")
    },
    warning = function(w) {
      invokeRestart("muffleWarning")
    }
  )
}

capture_plot_variant <- function(ob) {
  tmp_pdf <- tempfile(fileext = ".pdf")
  grDevices::pdf(tmp_pdf, width = 28, height = 6)
  on.exit({
    grDevices::dev.off()
    unlink(tmp_pdf)
  }, add = TRUE)
  quiet_eval(plot_variant(ob))
}

calc_unique_count <- function(x) {
  if (is.null(x) || !length(x)) {
    return(0L)
  }
  length(unique(x))
}

describe_filter_steps <- function(ob) {
  data.frame(
    Step = c(
      "1. redeemR.read.trim",
      "2. Create_redeemR_model",
      "3. clean_redeem",
      "4. add_annotation_redeem",
      "5. clean_redeem_remove_blacklist_RSRS50",
      "6. Add_DepthMatrix_filter2"
    ),
    Detail = c(
      sprintf(
        "thr=%s, edge_trim=%s, source=%s",
        unname(ob@para["Threhold"]),
        unname(ob@para["edge_trim"]),
        ob@attr$path
      ),
      sprintf(
        "qualifiedCellCut=%s, VAFcut=%s, Cellcut=%s",
        unname(ob@para["qualifiedCellCut"]),
        unname(ob@para["VAFcut"]),
        unname(ob@para["Cellcut"])
      ),
      "Binomial cleanup with fdr=0.05 and min_cell_per_variant=2",
      "Add population genetics and functional impact annotations",
      "Remove blacklist and RSRS50-associated calls from the active filtered state",
      "Attach filter2 depth matrix"
    ),
    stringsAsFactors = FALSE
  )
}

build_filtering_summary <- function(ob) {
  v_all <- as.data.frame(ob@V.fitered)
  gt_all <- as.data.frame(ob@GTsummary.filtered)
  variants_gtsummary <- as.data.frame(attr(ob, "GTsummary.ini"))
  homo_variants <- unique(ob@HomoVariants)
  qc_gt <- subset(gt_all, !Variants %in% c(homo_variants, hotcall_variants))
  qc_v <- subset(v_all, !Variants %in% hotcall_variants)

  data.frame(
    Metric = c(
      "VariantsGTSummary rows",
      "VariantsGTSummary unique cells",
      "VariantsGTSummary unique variants",
      "GTsummary.filtered rows",
      "GTsummary.filtered unique cells",
      "GTsummary.filtered unique variants",
      "V.fitered rows",
      "V.fitered heteroplasmic rows",
      "HomoVariants count",
      "QC variants after homo+hotcall exclusion",
      "QC cell-variant rows after homo+hotcall exclusion"
    ),
    Value = c(
      nrow(variants_gtsummary),
      calc_unique_count(variants_gtsummary$Cell),
      calc_unique_count(variants_gtsummary$Variants),
      nrow(gt_all),
      calc_unique_count(gt_all$Cell),
      calc_unique_count(gt_all$Variants),
      nrow(v_all),
      sum(v_all$HomoTag == "Hetero", na.rm = TRUE),
      length(homo_variants),
      calc_unique_count(qc_v$Variants),
      nrow(qc_gt)
    ),
    stringsAsFactors = FALSE
  )
}

choose_pct_column <- function(v_df) {
  pct_candidates <- c("PCTcell", "CellNPCT")
  pct_candidates[pct_candidates %in% colnames(v_df)][1]
}

top_variant_table <- function(v_df, n = 20) {
  if (!nrow(v_df)) {
    return(data.frame())
  }
  pct_col <- choose_pct_column(v_df)
  if (is.na(pct_col) || is.null(pct_col)) {
    pct_col <- "CellN"
  }
  out <- v_df %>%
    mutate(max_heteroplasmy = PositiveMean) %>%
    arrange(desc(.data[[pct_col]]), desc(CellN), desc(maxcts)) %>%
    dplyr::select(any_of(c(
      "Variants", pct_col, "CellN", "PositiveMean", "PositiveMean_cts",
      "maxcts", "totalVAF", "CV", "HomoTag", "changes", "types",
      "gene", "impact", "Disease"
    ))) %>%
    utils::head(n)
  names(out)[names(out) == pct_col] <- "PctCells"
  out
}

variant_count_bin_summary <- function(gt_df) {
  if (!nrow(gt_df)) {
    return(data.frame(Bin = character(), Count = integer(), Fraction = numeric()))
  }
  counts <- c(
    `1 UMI` = sum(gt_df$Freq == 1, na.rm = TRUE),
    `2 UMI` = sum(gt_df$Freq == 2, na.rm = TRUE),
    `>=3 UMI` = sum(gt_df$Freq >= 3, na.rm = TRUE)
  )
  total <- sum(counts)
  data.frame(
    Bin = names(counts),
    Count = unname(counts),
    Fraction = if (total > 0) round(unname(counts) / total, 4) else 0,
    stringsAsFactors = FALSE
  )
}

metrics_row <- function(sample_name, ob, qc_result, qc_status, qc_message) {
  gt_all <- as.data.frame(ob@GTsummary.filtered)
  v_all <- as.data.frame(ob@V.fitered)
  variants_gtsummary <- as.data.frame(attr(ob, "GTsummary.ini"))
  qc_gt <- subset(gt_all, !Variants %in% c(unique(ob@HomoVariants), hotcall_variants))
  pct_col <- choose_pct_column(v_all)
  data.frame(
    sample = sample_name,
    threshold = unname(ob@para["Threhold"]),
    variants_gtsummary_rows = nrow(variants_gtsummary),
    variants_gtsummary_unique_variants = calc_unique_count(variants_gtsummary$Variants),
    filtered_rows = nrow(gt_all),
    filtered_unique_variants = calc_unique_count(gt_all$Variants),
    v_filtered_rows = nrow(v_all),
    hetero_variants = sum(v_all$HomoTag == "Hetero", na.rm = TRUE),
    homo_variants = length(unique(ob@HomoVariants)),
    qc_matrix_rows = nrow(qc_gt),
    qc_matrix_variants = calc_unique_count(qc_gt$Variants),
    top_variant_pct_cells = if (!is.null(pct_col) && nrow(v_all)) max(v_all[[pct_col]], na.rm = TRUE) else NA_real_,
    qc_status = qc_status,
    qc_message = qc_message,
    stringsAsFactors = FALSE
  )
}

filter_steps <- describe_filter_steps(redeem_obj)
filter_summary <- build_filtering_summary(redeem_obj)
show_text <- capture_show(redeem_obj)
variants_preview <- safe_head(as.data.frame(attr(redeem_obj, "GTsummary.ini")), 3)
filtered_preview <- safe_head(redeem_obj@GTsummary.filtered, 10)
top_variants <- top_variant_table(as.data.frame(redeem_obj@V.fitered), params$top_n_variants)

qc_gtsummary <- subset(
  as.data.frame(redeem_obj@GTsummary.filtered),
  !Variants %in% c(unique(redeem_obj@HomoVariants), hotcall_variants)
)
vc_bins <- variant_count_bin_summary(qc_gtsummary)

qc_status <- "not_run"
qc_message <- ""
qc_result <- NULL
variant_plots <- NULL
expensive_qc_enabled <- nrow(qc_gtsummary) <= 30000L
expensive_qc_note <- if (expensive_qc_enabled) "" else sprintf("Skipped expensive plot/QC chunks for large object: %s GTsummary.filtered rows", nrow(qc_gtsummary))
```

## Metadata

```{r}
metadata_df <- data.frame(
  Field = c("Sample", "Threshold", "Input RDS", "redeemV final dir", "Report generated"),
  Value = c(
    sample_name,
    thr,
    normalizePath(params$input_rds),
    final_dir_abs,
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
  ),
  stringsAsFactors = FALSE
)
kable(metadata_df)
```

## Upstream redeemV QC

```{r}
if (file.exists(redeemv_qc_png_local)) {
  knitr::include_graphics(redeemv_qc_png_local)
} else {
  cat("Upstream redeemV QC PNG not found at:", redeemv_qc_png, "\n")
}
```

## Filtering Summary

```{r}
kable(filter_steps)
```

```{r}
kable(filter_summary)
```

## Object Summary

```{r}
cat("```text\n", show_text, "\n```\n", sep = "")
```

### VariantsGTSummary Preview

```{r}
kable(variants_preview)
```

### GTsummary.filtered Preview

```{r}
kable(filtered_preview)
```

## Depth QC

```{r, fig.width=14, fig.height=5, out.width="100%"}
depth_plots <- safe_plot(plot_depth(redeem_obj))
if (inherits(depth_plots, "error")) {
  cat("Depth QC failed:", conditionMessage(depth_plots), "\n")
} else {
  print(depth_plots$combined)
}
```

## Mutation Profile

```{r, fig.width=12, fig.height=5, out.width="100%"}
mutation_profile_plot <- safe_plot(MutationProfile.bulk(redeem_obj@UniqueV))
if (inherits(mutation_profile_plot, "error")) {
  cat("MutationProfile.bulk failed:", conditionMessage(mutation_profile_plot), "\n")
} else {
  print(mutation_profile_plot)
}
```

## Variant QC

### Top Variants by Percent Cells

```{r}
kable(top_variants)
```

### Cell-Variant Count Bins

```{r}
kable(vc_bins)
```

### Variant Plots

```{r}
if (!expensive_qc_enabled) {
  cat(expensive_qc_note, "\n")
} else {
  variant_plots <- tryCatch(
    capture_plot_variant(redeem_obj),
    error = function(e) e
  )
  if (inherits(variant_plots, "error")) {
    cat("plot_variant failed:", conditionMessage(variant_plots), "\n")
  }
}
```

#### Feature totalVAF versus coefficient of variation

```{r, fig.width=7, fig.height=5, out.width="100%"}
if (expensive_qc_enabled && !inherits(variant_plots, "error") && !is.null(variant_plots$p1)) {
  print(variant_plots$p1)
}
```

#### Heteroplasmy distribution

```{r, fig.width=7, fig.height=5, out.width="100%"}
if (expensive_qc_enabled && !inherits(variant_plots, "error") && !is.null(variant_plots$p2)) {
  print(variant_plots$p2)
}
```

#### Variant cell count versus max UMI count

```{r, fig.width=7, fig.height=5, out.width="100%"}
if (expensive_qc_enabled && !inherits(variant_plots, "error") && !is.null(variant_plots$p3)) {
  print(variant_plots$p3)
}
```

#### Variants per qualified cell

```{r, fig.width=7, fig.height=5, out.width="100%"}
if (expensive_qc_enabled && !inherits(variant_plots, "error") && !is.null(variant_plots$p4)) {
  print(variant_plots$p4)
}
```

## Consensus QC

```{r, fig.width=24, fig.height=7, out.width="100%"}
if (!expensive_qc_enabled) {
  cat(expensive_qc_note, "\n")
} else {
  consensus_error <- tryCatch(
    {
      quiet_eval(Show_Consensus(redeem_obj, N = params$consensus_n))
      NULL
    },
    error = function(e) e
  )
  if (inherits(consensus_error, "error")) {
    cat("Show_Consensus failed:", conditionMessage(consensus_error), "\n")
  }
}
```

## Filter2 QC

```{r}
if (!expensive_qc_enabled) {
  qc_status <- "skipped_large_object"
  qc_message <- expensive_qc_note
} else {
  redeem_obj_qc <- tryCatch(
    quiet_eval(add_raw_fragment(redeem_obj, raw = params$raw_name)),
    error = function(e) e
  )
  if (inherits(redeem_obj_qc, "error")) {
    qc_status <- "raw_fragment_failed"
    qc_message <- conditionMessage(redeem_obj_qc)
  } else {
  qc_attempt <- tryCatch(
    {
      grDevices::pdf(NULL)
      on.exit(grDevices::dev.off(), add = TRUE)
      quiet_eval(run_redeem_qc(redeem_obj_qc, redeem_obj_qc@HomoVariants))
    },
    error = function(e) e
  )
  if (inherits(qc_attempt, "error")) {
    qc_status <- "qc_failed_but_report_rendered"
    qc_message <- conditionMessage(qc_attempt)
  } else {
    qc_status <- "ok"
    qc_result <- qc_attempt
  }
  }
}
status_df <- data.frame(
  qc_status = qc_status,
  qc_message = ifelse(nzchar(qc_message), qc_message, ""),
  stringsAsFactors = FALSE
)
kable(status_df)
```

```{r}
if (!is.null(qc_result)) {
  kable(data.frame(
    Metric = names(qc_result$transversion_rate),
    Value = unname(qc_result$transversion_rate),
    stringsAsFactors = FALSE
  ))
}
```

```{r}
if (!is.null(qc_result)) {
  metric_block <- qc_result$report_metric
  metric_df <- data.frame(
    Metric = c(
      "number_of_total_mut",
      "number_of_total_cells",
      "median_mut_per_cell",
      "median_cells_connected",
      "fraction_in_component"
    ),
    Value = c(
      metric_block$number_of_total_mut,
      metric_block$number_of_total_cells,
      median(metric_block$mut_per_cell),
      median(metric_block$number_cell_connected),
      metric_block$fraction_in_component
    ),
    stringsAsFactors = FALSE
  )
  kable(metric_df)
}
```

```{r, fig.width=18, fig.height=5, out.width="100%"}
if (!is.null(qc_result)) {
  print(qc_result$plots$p_pos)
  print(qc_result$plots$pos_1mol)
  print(qc_result$plots$p_cell_maxcts)
  print(qc_result$plots$p_cell_meancts)
}
```

```{r}
metrics_df <- metrics_row(sample_name, redeem_obj, qc_result, qc_status, qc_message)
if (!is.null(params$output_metrics_tsv) && nzchar(params$output_metrics_tsv)) {
  write.table(metrics_df, params$output_metrics_tsv, sep = "\t", quote = FALSE, row.names = FALSE)
}
if (!is.null(params$output_report_rds) && nzchar(params$output_report_rds)) {
  saveRDS(
    list(
      sample_name = sample_name,
      threshold = thr,
      filtering_summary = filter_summary,
      variant_count_bins = vc_bins,
      qc_status = qc_status,
      qc_message = qc_message,
      qc_result = qc_result
    ),
    params$output_report_rds
  )
}
```
