---
title: "redeemR2.0 workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{redeemR2.0 workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`redeemR2.0` is the downstream analysis framework for processing REDEEM-V
mitochondrial consensus variant calls before lineage reconstruction. It imports
consensus-filtered calls from a REDEEM-V `final/` directory, applies the default
filter2 post-consensus filtering workflow, annotates variants, builds matched
cell-by-variant matrices, and produces a lineage-ready `redeemR` object.

The core workflow has five tasks:

1. Parse consensus-filtered mutation calls from REDEEM-V.
2. Apply post-consensus variant filtering and quality control.
3. Remove residual fragment-end artifacts with edge trimming.
4. Remove residual technical artifacts with binomial noise filtering.
5. Flag inherited or pre-existing variants and annotate functional impact.

In `redeemR2.0`, **filter2** refers to the default post-consensus filtering
workflow. It is applied after UMI-based consensus variant calling and is designed
to retain high-fidelity somatic mtDNA variants while removing residual technical
artifacts, inherited or pre-existing polymorphisms, and variants less suitable
for neutral lineage tracing.

## Inputs

The workflow starts from a REDEEM-V `final/` output directory. Unless otherwise
specified, use the stringent consensus output (`thr = "S"`).

The directory should contain:

```text
QualifiedTotalCts
RawGenotypes.Sensitive.StrandBalance
RawGenotypes.Specific.StrandBalance
RawGenotypes.Total.StrandBalance
RawGenotypes.VerySensitive.StrandBalance
```

The file roles are:

| File | Role |
|---|---|
| `RawGenotypes.*.StrandBalance` | Consensus-filtered mutation calls at a selected stringency |
| `QualifiedTotalCts` | Per-cell, per-position mtDNA depth used to build the matched depth matrix |

The supported thresholds are:

| Threshold | REDEEM-V file | Meaning |
|---|---|---|
| `T` | `RawGenotypes.Total.StrandBalance` | Total calls before consensus stringency filtering |
| `LS` | `RawGenotypes.VerySensitive.StrandBalance` | Less stringent consensus calls |
| `S` | `RawGenotypes.Sensitive.StrandBalance` | Stringent consensus calls; default |
| `VS` | `RawGenotypes.Specific.StrandBalance` | Very stringent consensus calls |

## Command-line quick start

For routine preprocessing, use the package script:

```bash
Rscript scripts/redeemR2.0_preprocess.R \
  --name sample1 \
  --input /path/to/redeemV/final \
  --output preprocessing/filter2/sample1/sample1.S.redeemR_filter2_adddepth.rds \
  --thr S \
  --edge-trim 9 \
  --min-variant-depth 5 \
  --do-median-depth-filter
```

This writes a cleaned `redeemR` object. The default project-local layout is:

```text
preprocessing/filter2/<sample>/<sample>.<thr>.redeemR_filter2_adddepth.rds
```

Run the separate QC report after preprocessing:

```bash
Rscript inst/scripts/render_filter2_qc.R \
  --input-rds preprocessing/filter2/sample1/sample1.S.redeemR_filter2_adddepth.rds \
  --output-html reports/filter2_qc/sample1/sample1.S.filter2_qc.html \
  --sample-name sample1 \
  --thr S \
  --output-metrics-tsv preprocessing/filter2/sample1/sample1.S.filter2_qc_metrics.tsv \
  --output-report-rds preprocessing/filter2/sample1/sample1.S.filter2_qc_report.rds
```

See `vignettes/filter2-qc.Rmd` for a detailed description of the QC report.

## R workflow

Load the package and point to the REDEEM-V `final/` directory:

```{r setup}
library(redeemR)

redeemv_final <- "/path/to/redeemV/final"
sample_name <- "sample1"
thr <- "S"
edge_trim <- 9
```

### Parse and edge-trim REDEEM-V calls

`redeemR.read.trim()` reads the selected consensus genotype file, annotates raw
calls by distance to the nearest fragment end, removes calls within the trimmed
edge window, and reconstructs genotype summaries after trimming.

```{r parse-trim}
variants <- redeemR.read.trim(
  path = redeemv_final,
  thr = thr,
  edge_trim = edge_trim
)
```

The returned genotype summary stores per-cell variant UMI counts, site depth, and
heteroplasmy for retained cell-variant observations. It also carries attributes
used downstream, including the selected threshold, depth summaries, input path,
and edge-trim setting.

### Create the initial redeemR object

`Create_redeemR_model()` initializes the `redeemR` object, keeps cells with mean
mitochondrial coverage above the coverage threshold, identifies candidate
variants, constructs count and binary matrices, and runs the per-variant binomial
noise model.

```{r create-object}
obj <- Create_redeemR_model(
  variants,
  qualifiedCellCut = 10,
  VAFcut = 1,
  Cellcut = 2
)
```

At this point, the object contains initial genotype summaries, candidate variant
summaries, cell metadata, depth summaries, homoplasmic or near-homoplasmic
variant labels, and preliminary cell-by-variant matrices.

### Apply binomial post-consensus filtering

The binomial model compares each variant's UMI-count distribution across cells
with a single-binomial technical-noise model parameterized by the observed total
variant counts and position-specific mean coverage. P-values are converted to
q-values, and filter2 keeps heteroplasmic variants passing the FDR threshold and
minimum cell-count filter.

```{r clean-binomial}
obj <- clean_redeem(
  obj,
  fdr = 0.05,
  min_cell_per_variant = 2
)
```

### Annotate variants

Add population, sequence-context, and functional annotations to the filtered
variant table:

```{r annotate}
obj <- add_annotation_redeem(obj)
```

These annotations include population features, RSRS50 status, haplogroup marker
information, blacklist regions, homopolymer context, hypermutable-site labels,
amino-acid consequences, mitochondrial disease annotations, and
transition/transversion class when the corresponding annotation functions are
available.

### Remove RSRS50 and blacklist variants

For downstream lineage analysis, filter2 removes heteroplasmic RSRS50 variants
and variants falling in predefined mitochondrial blacklist regions.

```{r remove-rsrs50-blacklist}
obj <- clean_redeem_remove_blacklist_RSRS50(obj)
```

### Add the matched depth matrix

Build the cell-by-variant depth matrix from `QualifiedTotalCts` and match it to
the filtered count matrix.

```{r add-depth}
obj <- Add_DepthMatrix_filter2(obj)
```

The resulting matrix is stored in `obj@Ctx.Mtx.depth` and should have the same
cell and variant dimensions as the filtered count matrix.

### Add depth and UMI-support metrics

Add per-variant median depth and the number of cells with nonzero coverage:

```{r add-depth-metrics}
obj <- add_median_depth_to_redeemR(obj)
```

If desired, remove variants with low median depth. The default filter2 cutoff is
5 when this step is enabled.

```{r median-depth-filter}
obj <- clean_redeem_remove_low_median_depth(
  obj,
  min_median_depth = 5
)
```

Finally, add UMI-support features used during QC, including the fraction of
variant detections supported by two or three UMIs.

```{r umi-support}
obj <- add_prop_2_3_to_redeemR(obj)
```

### Save the filter2 object

```{r save-object}
out_rds <- file.path(
  "preprocessing/filter2",
  sample_name,
  paste0(sample_name, ".", thr, ".redeemR_filter2_adddepth.rds")
)

dir.create(dirname(out_rds), recursive = TRUE, showWarnings = FALSE)
saveRDS(obj, out_rds)
```

## Inspect the output object

The final object contains the filtered genotype records, variant annotations,
count matrices, binary matrices, depth matrix, and QC-ready metadata.

```{r inspect}
show(obj)
print_redeemR_matrix_dims(obj)
head(obj@V.fitered)
head(obj@GTsummary.filtered)
```

Important slots are:

| Slot | Description |
|---|---|
| `@V.fitered` | Filtered variant-level summary and annotations |
| `@GTsummary.filtered` | Filtered cell-variant genotype records |
| `@Cts.Mtx` | Cell-by-variant mutant allele count matrix |
| `@Cts.Mtx.bi` | Binarized cell-by-variant mutation matrix |
| `@Ctx.Mtx.depth` | Matched cell-by-variant depth matrix |
| `@HomoVariants` | Homoplasmic or near-homoplasmic variants identified during preprocessing |

## Basic QC plots

The following functions provide quick checks before formal QC report rendering:

```{r basic-qc}
plot_depth(obj)$combined
MutationProfile.bulk(obj@UniqueV)
variant_plots <- plot_variant(obj)
variant_plots$p1
variant_plots$p2
variant_plots$p3
variant_plots$p4
```

For the full report, render the filter2 QC report described in
`vignettes/filter2-qc.Rmd`.

## Default parameters

| Step | Default |
|---|---|
| Consensus input | Stringent consensus output, `thr = "S"` |
| Edge trimming | 9 bp from fragment ends |
| Cell coverage filter | Mean mtDNA coverage >= 10 |
| Initial variant cell filter | Detected in >= 2 cells |
| Supporting UMI threshold | At least 1 supporting UMI in an individual cell |
| Homoplasmic or near-homoplasmic annotation | `CellNPCT > 0.75`, `PositiveMean > 0.75`, `CV < 0.01` |
| Binomial filter | Per-variant goodness-of-fit test |
| FDR threshold | `q <= 0.05` |
| Optional median-depth filter | Remove variants with median depth < 5 |
| Downstream lineage exclusions | RSRS50 variants and mitochondrial blacklist regions |

## Downstream analysis

The filter2 object is designed to support downstream lineage reconstruction,
clonal analysis, and integration with single-cell multiome data. Typical next
steps include distance calculation, graph construction, tree reconstruction, and
clone or clade-level analysis using the filtered mutation matrices and matched
depth matrix.
