---
title: "Variant annotation in redeemR2.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variant annotation in redeemR2.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Variant annotation is a core part of the `redeemR2.0` filter2 workflow. After
post-consensus filtering, retained mtDNA variants are annotated with population,
sequence-context, functional, disease, and mutation-class features. These fields
support two goals:

1. identify variants that should be removed or flagged before neutral lineage
   tracing, such as inherited, pre-existing, or technically problematic variants;
2. provide interpretable biological context for the filtered variant set.

In the default filter2 preprocessing workflow, annotation is applied after
binomial filtering and before RSRS50/blacklist cleanup:

```{r annotation-position}
obj <- clean_redeem(obj, fdr = 0.05, min_cell_per_variant = 2)
obj <- add_annotation_redeem(obj)
obj <- clean_redeem_remove_blacklist_RSRS50(obj)
```

## Input table

Most annotation functions operate on a variant-level table with a `Variants`
column. Variant IDs use the standard redeemR format:

```text
<position>_<reference_base>_<alternate_base>
```

For example:

```{r example-input}
variant_annotation <- data.frame(
  Variants = c("73_A_G", "310_T_C", "11719_G_A", "14766_C_T"),
  CellN = c(5, 12, 100, 80),
  stringsAsFactors = FALSE
)
```

For a filter2 object, the active variant-level table is stored in:

```{r get-vfiltered}
variant_annotation <- obj@V.fitered
```

## One-line annotation

The simplest public entry point is `annotate_all_variants()`:

```{r annotate-all}
obj@V.fitered <- annotate_all_variants(obj@V.fitered)
```

For a `redeemR` object, the filter2 workflow usually calls the object wrapper:

```{r object-wrapper}
obj <- add_annotation_redeem(obj)
```

`add_annotation_redeem()` applies `annotate_all_variants()` to `obj@V.fitered`
and returns the updated object.

## Annotation components

`annotate_all_variants()` applies the annotation suite in this order:

```{r annotation-suite}
variant_annotation <- variant_annotation |>
  annotate_variants_hypermutable() |>
  annotate_variants_population_stats() |>
  annotate_variants_blacklist() |>
  annotate_variants_homopolymer() |>
  annotate_variants_aachange() |>
  annotate_variants_mito_disease() |>
  Annotate_base_change()
```

### Hypermutable-site labels

`annotate_variants_hypermutable()` labels variants that fall into the
hypermutable category using the package's population recurrence data.

```{r hypermutable}
variant_annotation <- annotate_variants_hypermutable(variant_annotation)
```

Main output column:

| Column | Meaning |
|---|---|
| `hyper_label` | `"hyper"` for variants identified as hypermutable; otherwise empty |

### Population and haplogroup statistics

`annotate_variants_population_stats()` adds MITOMAP population-frequency fields,
RSRS50 status, and haplogroup marker counts.

```{r population}
variant_annotation <- annotate_variants_population_stats(variant_annotation)
```

Main output columns:

| Column | Meaning |
|---|---|
| `Locus` | Mitochondrial feature or gene annotation from the population table |
| `RSRS50` | RSRS50 ancestral-state status |
| `freq_all` | Overall MITOMAP population frequency |
| `freq_african` | Frequency in lineage L |
| `freq_asian` | Frequency in lineage M |
| `freq_eurasian` | Frequency in lineage N |
| `n_haplos` | Number of haplogroup marker entries for the variant |

In filter2, heteroplasmic RSRS50 variants are removed before downstream lineage
analysis because they are more consistent with inherited or pre-existing ancestry
signals than neutral somatic lineage markers.

### Blacklist-region annotation

`annotate_variants_blacklist()` flags variants in known problematic mtDNA
regions prone to misalignment or ambiguous mapping.

```{r blacklist}
variant_annotation <- annotate_variants_blacklist(variant_annotation)
```

The blacklist intervals are:

| rCRS positions | Rationale |
|---|---|
| 302-315 | Misalignment-prone poly-C region |
| 513-525 | Misalignment-prone repetitive region |
| 3105-3109 | rCRS 3107N ambiguity region |
| 16182-16187 | Misalignment-prone poly-C region |

Main output column:

| Column | Meaning |
|---|---|
| `blacklist_region` | `"blacklist_region"` when a variant falls in a predefined blacklist interval |

Filter2 removes variants in these blacklist regions before downstream lineage
analysis.

### Homopolymer context

`annotate_variants_homopolymer()` labels whether a variant falls in a predefined
mitochondrial homopolymer interval.

```{r homopolymer}
variant_annotation <- annotate_variants_homopolymer(variant_annotation)
```

Main output column:

| Column | Meaning |
|---|---|
| `homopolymer` | Homopolymer region label, or `NA` if the variant is outside known intervals |

### Amino-acid consequence and predicted impact

`annotate_variants_aachange()` uses a mitochondrial coding reference and
`dndscv` to annotate coding variants.

```{r aachange}
variant_annotation <- annotate_variants_aachange(variant_annotation)
```

Main output columns:

| Column | Meaning |
|---|---|
| `gene` | Mitochondrial coding gene |
| `aachange` | Amino-acid change notation |
| `impact` | Predicted coding impact category |

This step requires the `dndscv` package and the mitochondrial coding reference
bundled with `redeemR`.

### Mitochondrial disease annotations

`annotate_variants_mito_disease()` joins known mitochondrial disease associations
from the package disease table.

```{r disease}
variant_annotation <- annotate_variants_mito_disease(variant_annotation)
```

Main output column:

| Column | Meaning |
|---|---|
| `Disease` | Known mitochondrial disease association, if available |

### Base-change and transition/transversion class

`Annotate_base_change()` adds nucleotide substitution and mutation-class fields.

```{r base-change}
variant_annotation <- Annotate_base_change(variant_annotation)
```

Main output columns:

| Column | Meaning |
|---|---|
| `changes` | Nucleotide substitution, such as `C_T` or `A_C` |
| `types` | `transition` or `transversion` |

These fields are used in filter2 QC to summarize transversion burden and to color
variant-support diagnostic plots.

## Cleaning annotated variants for lineage analysis

After annotation, the default filter2 workflow removes heteroplasmic RSRS50
variants and variants in mitochondrial blacklist regions:

```{r clean-annotated}
obj <- clean_redeem_remove_blacklist_RSRS50(obj)
```

This function updates the active variant table, filtered genotype summary, and
cell-by-variant matrices so downstream lineage reconstruction uses the cleaned
variant set.

## Recomputing depth and support annotations

After filtering and annotation cleanup, filter2 builds a matched depth matrix and
adds depth/support metrics:

```{r depth-support}
obj <- Add_DepthMatrix_filter2(obj)
obj <- add_median_depth_to_redeemR(obj)
obj <- clean_redeem_remove_low_median_depth(obj, min_median_depth = 5)
obj <- add_prop_2_3_to_redeemR(obj)
```

Important added columns include:

| Column | Meaning |
|---|---|
| `median_depth` | Median site depth across cells for the variant |
| `cellN_depth_gt0` | Number of cells with nonzero depth at the variant site |
| `CellNPCT` | Fraction of covered cells carrying the variant |
| `prop_2_3` | Fraction of detections supported by two or three UMIs |
| `mean_umi_gt1` | Mean UMI count among detections with more than one UMI |

When median-depth filtering is enabled, variants with median depth below 5 are
removed by default.

## Inspecting annotation output

Useful checks after annotation include:

```{r inspect-annotations}
head(obj@V.fitered)

table(obj@V.fitered$blacklist_region, useNA = "ifany")
table(obj@V.fitered$RSRS50, useNA = "ifany")
table(obj@V.fitered$types, useNA = "ifany")
table(obj@V.fitered$hyper_label, useNA = "ifany")
```

For downstream summaries, it is often useful to export the final annotated
variant table:

```{r export-annotations}
write.csv(
  obj@V.fitered,
  file = "sample1.filter2.variant_annotations.csv",
  row.names = FALSE
)
```

## Recommended reporting fields

For manuscript or project-level QC summaries, the most useful annotation fields
are usually:

- `Variants`
- `CellN`
- `CellNPCT`
- `PositiveMean`
- `median_depth`
- `prop_2_3`
- `types`
- `RSRS50`
- `blacklist_region`
- `homopolymer`
- `hyper_label`
- `freq_all`
- `n_haplos`
- `gene`
- `aachange`
- `impact`
- `Disease`

These columns document both why variants were retained or removed and what
biological context is available for the retained set.
